---
title: "SAS and SPSS"
---

LISSY supports SAS (9.4) and SPSS (22) in addition to R and Stata. 

Note that the user guide does not provide full coverage of SAS and SPSS workflows. The examples below are intended as a starting points.


## SAS

Data is loaded by referencing the dataset macro provided by LISSY. The example below computes basic summary statistics on the Luxembourg 2010 household file.


```{r, eval =FALSE, echo=TRUE}
PROC MEANS DATA=&lu10h;
```


To produce and export a chart, SAS requires specifying the output file via a `FILENAME` statement before calling the graphics procedure. The `GOPTIONS` line sets the output device and format. Without this, no image will be returned by LISSY.


```{r, eval =FALSE, echo=TRUE}
FILENAME output "&mypdf/graphtestsas.png";
GOPTIONS RESET=ALL DEVICE=PNG GSFNAME=output GSFMODE=REPLACE;
PROC GCHART data=&lu10p;
VBAR age / TYPE=percent;
RUN;
```



## SPSS 

In SPSS, datasets are loaded using `GET FILE` with the LISSY-provided dataset macro. The example below loads the Luxembourg 2010 household file.



```{r, eval =FALSE, echo=TRUE}
get file = lu10h
```


To produce a chart and export it, combine the `GRAPH` command with `OUTPUT EXPORT`. The example below produces a bar chart of relationship status by years of education and saves it as a JPEG.


```{r, eval =FALSE, echo=TRUE}
GET FILE = lu10p
GRAPH /BAR(SIMPLE)=relation BY edyrs.
OUTPUT EXPORT /JPG DOCUMENTFILE="mypdf\graphtestspss.jpg
```