---
title: "Describing Data"
---


# Continuous Variables


In LIS and LWS datasets, continuous variables are distinct from the socioeconomic categorical variables discussed in the previous section. Rather than grouping households or individuals into discrete classes, they capture actual flows and stocks — income, consumption, and wealth, along with their subcomponents and aggregates. Examples include disposable household income (`dhi`), capital income (`hicapital`), and net worth (`dnw`). These variables tend to be right-skewed, can span several orders of magnitude, and vary considerably in coverage across country-years — a variable that is well-populated in one dataset may be sparse or entirely absent in another. Getting a quick distributional snapshot before any substantive analysis is therefore good practice. The emphasis in this section is not so much on the mechanics of any single statistic, but on building reproducible workflows that apply consistently across countries and waves with minimal extra effort.


<br>

## Quick Summary

A natural starting point is to combine the base function `summary()` with the familiar `purrr::map()` function to get an initial sense of a variable's scale and distribution.

::: {.callout-tip}
Always make sure the data is at the appropriate level for the variable being analysed. For instance, running summary statistics for disposable household income (`dhi`), defined at the household-level, from an individual-level dataset will silently duplicate observations, inflating counts and distorting distributional measures.
:::


```{r, eval=TRUE, echo=TRUE, results = 'hold', message = FALSE, warning = FALSE}
library(lissyrtools)
library(dplyr)
library(purrr)

it_data = lissyuse(data = 'it', vars = 'dhi', from = 2005)

it_data %>% purrr::map(~ summary(.x$dhi))
```

<div style="height: 30px;"></div>


::: {.callout-warning}
While `summary()` provides a convenient first look, its distributional measures (e.g., the median, p25, and p75) **do not follow the methodology used by LIS** in its official figures. This is discussed in more detail [later in this section](/r-users/dd_cont_vars.html#pm).
:::



<br>

## Mapping a Descriptive Function

Beyond summary statistics, it is often useful to build small helper functions that capture a specific aspect of a variable and then apply them systematically across all available datasets. This approach keeps the workflow modular and easy to extend.

As a practical illustration, consider measuring data coverage: for a given income component, what share of observations are both non-missing and non-zero? This is especially relevant for variables like capital income (`hicapital`), which may be consistently recorded in some country-years but sparse or absent in others.


```{r, eval=TRUE, echo=TRUE, results = 'hold', message = FALSE, warning = FALSE, class.output = "foldable"}
it_data = lissyuse(data = 'it', vars = c('dhi', 'hicapital'), from = 2005)

share_non_missing_non_zero <- function(df, variable_of_interest) {

    denominator = nrow(df)
    numerator   = sum( (!is.na(df[variable_of_interest])) & (df[variable_of_interest] !=0) )
    numerator_zero = sum((df[variable_of_interest] == 0)) 

    share_to_out = ( (numerator) / denominator ) * 100
    return(share_to_out)
}

purrr::map(it_data, ~ share_non_missing_non_zero(.x, 'hicapital'))

```

<div style="height: 30px;"></div>

The result immediately reveals whether a given income component is consistently observed across the selected waves, or whether its coverage varies substantially over time, making it a useful diagnostic before including the variable in any analysis.

<br>

## Subgroup descriptives 

Aggregate statistics can conceal important variation across groups. For continuous variables, computing descriptives by subgroup makes it possible to compare, for instance, how disposable income differs across tenure statuses or demographic categories.

The function `run_weighted_mean()` handles this directly via its `by` argument, applying the same logic consistently across all loaded datasets:

```{r, eval=TRUE, echo=TRUE, results = 'hold', message = FALSE, warning = FALSE, class.output = "foldable"}
ex_data = lissyuse(data = c('fr', 'it', 'es', 'de'), , vars = c('dhi', 'own'), from = 2021)

run_weighted_mean(ex_data, var_name = 'dhi', wgt_name = NULL,  by = 'own')
```

<div style="height: 30px;"></div>

Like `run_weighted_count()` introduced in the previous section, `run_weighted_mean()` makes subgroup descriptives straightforward to produce in a reproducible and consistent way across multiple country-year datasets.

<!-- Make sure this line here makes sense or if it was introduced repeatedly many time in the sections before -->
For further processing, users may wish to transform these outputs into a tidy tabular format suitable for export or visualisation. Guidance on structuring and exporting results is available in the [dedicated export section](/r-users/tidy_exp_to_csv.html#export-csv).


<!-- add perhaps something here on `inum` and LWS !!!!!!!-->

<br>

## Percentiles methodology {#pm}

We close with an important note on how percentiles are estimated, as the choice of method directly affects the comparability of results with LIS official figures.

LIS adopts Definition 4 from Hyndman and Fan (1996) for the computation of all percentile-based estimates, including medians, percentile ratios, poverty lines, and decile shares. Full details are available in this [LIS methodological note](https://www.lisdatacenter.org/news-and-events/note-on-percentile-methodology-used-in-lis-key-figures/).

The function `run_weighted_percentiles()` follows this methodology by default, so users can rely on it to produce estimates that are fully consistent with LIS standards without needing to configure anything manually.


<!--While this level of detail may go beyond the needs of an initial data exploration, being aware of the underlying methodology is important. Using a consistent and well-defined approach to percentile estimation helps avoid discrepancies that can arise when combining results from different functions or packages.-->


```{r, eval=TRUE, echo=TRUE, results = 'hold', message = FALSE, warning = FALSE}
spain_data <- lissyuse(data = "es", vars = "dhi", from = 2020)

run_weighted_percentiles(spain_data, var_name = 'dhi', probs = 0.5)
```

<div style="height: 30px;"></div>

The `probs` argument accepts any value or vector of values between 0 and 1, making it straightforward to retrieve the full set of percentiles in a single call:

```{r, eval=TRUE, echo=TRUE, results = 'hold', message = FALSE, warning = FALSE}    
run_weighted_percentiles(spain_data, var_name = 'dhi', probs = seq(0.1, 0.9, 0.1), share = FALSE, average = FALSE)
```

<div style="height: 30px;"></div>

Beyond the point estimates themselves, the function also supports the computation of distributional shares, that is, the share of total income held by each segment of the distribution:


```{r, eval=TRUE, echo=TRUE, results = 'hold', message = FALSE, warning = FALSE}
run_weighted_percentiles(spain_data, var_name = 'dhi', probs = seq(0, 1, 0.1), share = TRUE, average = FALSE)
```

<div style="height: 30px;"></div>

And the average income within each group:

```{r, eval=TRUE, echo=TRUE, results = 'hold', message = FALSE, warning = FALSE}
run_weighted_percentiles(spain_data, var_name = 'dhi', probs = seq(0, 1, 0.1), average = TRUE, share = FALSE) 
```


