---
title: "Data Handling"
---
# Processing Income Variables

Due to the wide range of research questions that can be explored using data from LIS, it is often necessary to perform multiple and distinct types of data preparation after the initial import. For this reason, this section is divided into two parts, each dedicated to a different analytical goal.

In the first part, we demonstrate how to compute a Children Poverty Rate among Single-Mother Families. This particular indicator requires several steps, making it a comprehensive example of the types of functions users can rely on, as well as highlighting important considerations when working with LIS and LWS data.

In the second part, we focus on how to apply the deflators provided by LIS, an essential step for cross-country and cross-time comparisons, and discuss the specific nuances involved in that process.

<br>

## Children Poverty Rates - Single-Mother Families

The sections that follow illustrate a more comprehensive workflow, combining several preparation steps to construct a policy-relevant indicator from microdata.

The aim will be to compute the poverty rate among children in single-mother households in several countries. Constructing this indicator requires combining information available at both the household level (`dhi`, `nhhmem17`, `hpartner`), and the individual level (`sex`).

::: callout-important
Recall from the earlier sections of this guidebook when we introduced `lissyuse()`, that when both household and individual level variables  are specified in `vars`, both files for a given `ccyy` are loaded and merged automatically. As a result, the data is loaded at the individual level.
:::

```{r, warning = FALSE, message = 5}
library(lissyrtools)
library(dplyr)
library(purrr)

data <- lissyuse(
  data = c("us", "de", "uk", "fr"),  
  vars = c("dhi", "nhhmem17", "sex", "hpartner"), # 3 'h-level' + 1 'p-level'
  from = 2018, 
  to = 2024   
)  
```



```{r, warning = FALSE, message = FALSE, class.output = "foldable"}  
columns = names(data[[1]]) 
lissyrtools::variable_labels(vars = columns)
```


::: callout-note
Variables `relation` and `nhhmem` are part of the group of variables that are always imported by default.
:::

<br> 

## Filtering and Variable Construction

At this stage, we prepare the dataset so that it is **aligned with the unit of analysis required for our indicator**: the household. Although we imported both household- and person-level variables, the poverty measure is computed at the household level and then attributed to individuals (in this case, children).

We define a pipeline inside `purrr::map()` to apply the same sequence of transformations
to each dataset. We begin by retaining only records corresponding to the household
reference person (`relation == 1000`), ensuring that each household appears exactly once
in the dataset and avoiding the duplication that would arise from keeping all person-level
records. We then filter out observations with missing disposable household income (`dhi`),
as these cannot contribute to the poverty calculation.

```{r}
data_prep <- data %>%
  purrr::map(~ .x %>% 
        filter(relation == 1000) %>% 
        filter(!is.na(dhi)) %>% 
        mutate(
          new_wgt = hwgt * nhhmem,
          new_chld_wgt = hwgt * nhhmem17,
        )
  )
```


We then construct adjusted weights. For population-level indicators, we use a person-level weight derived from the household weight multiplied by household size (`hwgt * nhhmem`). For child-specific indicators, we instead construct a child weight by multiplying the household weight by the number of children in the household (`hwgt * nhhmem17`). These weights ensure that estimates correctly reflect the underlying population structure.

<br>

## Bottom and Top Coding

Before proceeding with inequality and poverty calculations, we address the presence of extreme income values. While the underlying microdata do not impose bounds on income, very high or very low values can disproportionately influence distributional measures.

To mitigate this, we apply bottom- and top-coding to disposable household income (dhi), with the help of an auxilliary function named `apply_iqr_top_bottom_coding()` tailored specifically for this step.

```{r}
data_tb_coded <- data_prep %>%
    lissyrtools::apply_iqr_top_bottom_coding(
        var_name = 'dhi', 
        wgt_name = 'hwgt', 
        times = 3, 
        one_sided = NULL,
        type = 'type_4' 
        )
```

This adjustment is carried out on the logarithmic scale using an interquartile range (IQR) rule: values above the upper threshold (Q3 + 3×IQR) are capped, and values below the lower threshold (Q1 − 3×IQR) are raised. This method ensures that extreme observations do not disproportionately affect the results, while preserving the overall shape and structure of the income distribution.

::: callout-note
To ensure consistency with the methodology applied by LIS across all disseminated aggregate figures, this step must be performed prior to equivalisation. In addition, the quartiles used to define the IQR thresholds should be estimated following Definition 4 as proposed by Hyndman and Fan (1996), which corresponds to the `type = "type_4"` argument above
:::

<!-- DART still uses old percentile estimation method in its IQR preparation... -->

<br>

## Equivalence Scales

Households differ in size and composition, which affects their economic needs. To enable meaningful comparisons across households, we equivalize disposable household income.

We apply the square root equivalence scale, dividing household income by the square root of the total number of household members. This transformation accounts for economies of scale in consumption: larger households require more resources, but not proportionally more per additional member.

As a result, all individuals within a household are assigned the same equivalised income, regardless of their age or role within the household.

```{r}
data_eqv <- data_tb_coded %>%
     lissyrtools::apply_sqrt_equivalisation(
        var_name = 'dhi'
        )
```

<br>

## Poverty Line

Having adjusted the income variable and reduced the dataset to the household level, we are now ready to compute the poverty line. We define it as **50% of the weighted median of equivalised income**, which is the standard relative poverty threshold used in cross-national research.

We can use the already familiar `run_weighted_percentiles()`, which by default computes the minimum, maximum, and the 25th, 50th, and 75th percentiles, returning them as a named vector inside a list. The call below previews the output structure for the first dataset.

```{r,  class.output = "foldable"}
lissyrtools::run_weighted_percentiles(
        data_eqv[1], 
        var_name = "dhi", 
        wgt_name = "new_wgt"
    )
```

::: {style="height: 30px;"}
:::

We then use `purrr::map()` to apply this function across all datasets, retrieve the element corresponding to the median (named `"50%"`), and multiply it by 0.5 to establish the poverty line. This relative definition ensures that poverty is assessed in relation to the living standards prevailing within each dataset, so there is no need to adjust for inflation or different currencies.

```{r,  class.output = "foldable"}
medians <- purrr::map(
    lissyrtools::run_weighted_percentiles(
        data_eqv, 
        var_name = "dhi", 
        wgt_name = "new_wgt"
    ), "50%"
)
poverty_lines <- purrr::map(medians  , ~ .x  * 0.5)

print(poverty_lines)
```


::: {style="height: 30px;"}
:::

It is important to note that the poverty line is always calculated based on the **full population**. This same threshold is subsequently applied when evaluating poverty rates for specific subgroups, such as children in single-mother households.

<br>

## Conditional Variable Construction

With the poverty line established, we create an indicator identifying whether each household falls below this threshold. For each observation, we compare equivalised income (`dhi`) to the poverty line.

Households with income below the threshold are classified as poor and assigned a value of 100, all others are assigned 0.

```{r}
data_with_pov_line <- purrr::map2(
  data_eqv,
  poverty_lines,
  ~ .x %>%
    dplyr::mutate(
      line = .y,
      poor = dplyr::if_else(dhi < line, 100, 0)
    )
)
```

::: callout-tip
When using `purrr::map2()`, `.x` refers to the current element of the first list (`data_eqv`), while `.y` refers to the corresponding element of the second list (`poverty_lines`). This allows both lists to be iterated over in lockstep.
:::

<br>

## Weighted Aggregation

In the final step, we restrict the sample to the population of interest: children living in single-mother households. This is defined as households where the reference person is female (`sex == 2`), has no partner present (`hpartner == 0`), and where at least one child is present (`nhhmem17 > 0`).

We then compute the weighted mean of the poverty indicator using the child-specific weight (`hwgt * nhhmem17 = new_chld_wgt`). This produces the poverty rate among children in single-mother households, expressed as a percentage.

Throughout this process, the poverty line remains fixed, having been calculated for the total population, while the weighting scheme ensures that the estimate accurately reflects the distribution of children across households.


```{r}
data_with_pov_line %>% 
  purrr::map( ~ .x %>%
         filter(hpartner==0 & sex==2 & nhhmem17>0)
  ) %>% 
  run_weighted_mean("poor", "new_chld_wgt")
```





