---
title: "Data Handling"
---
## Preparing Income Variables

Due to the wide range of research questions that can be explored using data from LIS, it is often necessary to perform multiple and distinct types of data preparation after the initial import. For this reason, this section is divided into two parts, each dedicated to a different analytical goal.

In the first part, we demonstrate how to compute a Children Poverty Rate among Single-Mother Families. This particular indicator requires several steps, making it a comprehensive example of the types of R functions users can rely on, as well as highlighting important considerations when working with LIS and LWS data.

Secondly, to work effectively on cross-country projects with LIS data, users must have a solid understanding of how to apply the deflators provided. Therefore, the second part will therefore focus on this process and its specific nuances.

### Children Poverty Rates - Single-Mother Families

The sections that follow illustrates a more comprehensive workflow, combining several preparation steps to construct a policy-relevant indicator from microdata.

The aim will be to compute the poverty rate among children in single-mother households in the United States. Constructing this indicator requires combining information available at both the household level (`dhi`, `nhhmem17`, `hpartner`), and the individual level (`sex`).

::: callout-important
Recall from the early sections of this guidebook when we introduced `lissyuse()`, that when both household and individual levels are specified in `vars`, both files for a given `ccyy` are loaded are merged automatically. Therefore we end up with the data loaded at the individual level.
:::

```{r, warning = FALSE, message = 5}
library(lissyrtools)
library(dplyr)
library(purrr)

us_data <- lissyuse(
  data = "us",  
  vars = c("dhi", "nhhmem17", "sex", "hpartner"), # 3 'h-level' + 1 'p-level'
  from = 2018, 
  to = 2024   
)  
```



```{r, warning = FALSE, message = FALSE, class.output = "foldable"}  
columns = names(us_data[[1]]) 
lissyrtools::variable_labels(vars = columns)
```


::: callout-note
Variables `relation` and `nhhmem` are part of the group of variables that are always imported by default.
:::


### Filtering and Variable Construction

At this stage, we prepare the dataset so that it is **aligned with the unit of analysis required for our indicator**: the household. Although we imported both household- and person-level variables, the poverty measure is computed at the household level and then attributed to individuals (in this case, children).

```{r}
data_prep <- us_data %>%
  purrr::map(~ .x %>% 
        filter(!is.na(dhi)) %>% 
        filter(relation == 1000) %>% 
        mutate(
          new_wgt = hwgt * nhhmem,
          new_chld_wgt = hwgt * nhhmem17,
        )
  )
```

Like in previous examples, we will define a pipeline inside the `purrr::map()` function to apply the same processes for each dataset. We begin by filtering out observations with missing disposable household income (`dhi`). Next, we retain only records corresponding to the household reference person (`relation == 1000`). This step ensures that each household appears only once in the dataset, avoiding duplication that would arise from person-level records.

We then construct adjusted weights. For population-level indicators, we use a person-level weight derived from the household weight multiplied by household size (`hwgt * nhhmem`). For child-specific indicators, we instead construct a child weight by multiplying the household weight by the number of children in the household (`hwgt * nhhmem17`). These weights ensure that estimates correctly reflect the underlying population structure.

### Bottom and Top Coding

Before proceeding with inequality and poverty calculations, we address the presence of extreme income values. While the underlying microdata do not impose bounds on income, very high or very low values can disproportionately influence distributional measures.

To mitigate this, we apply bottom- and top-coding to disposable household income (dhi), with the help of another auxilliary function named `apply_iqr_top_bottom_coding()` tailored specifically for such stage.

```{r}
data_tb_coded <- data_prep %>%
    lissyrtools::apply_iqr_top_bottom_coding(
        var_name = 'dhi', 
        wgt_name = 'hwgt', 
        times = 3, 
        one_sided = NULL,
        type = 'type_4' 
        )
```

In more detail, this adjustment is carried out on the logarithmic scale using an interquartile range (IQR) rule: values above the upper threshold (Q3 + 3×IQR) are capped, and values below the lower threshold (Q1 − 3×IQR) are raised. This method ensures that extreme observations do not disproportionately affect the results, while preserving the overall shape and structure of the income distribution.

::: callout-note
To ensure consistency with the latest methodology applied by LIS across all disseminated aggregate figures, this step must be performed prior to equivalisation. In addition, the quartiles used to define the IQR thresholds should be estimated following Definition 4 as proposed by Hyndman and Fan (1996).
:::

<!-- DART still uses old percentile estimation method in its IQR preparation... -->

### Equivalence Scales

Households differ in size and composition, which affects their economic needs. To enable meaningful comparisons across households, we equivalize disposable household income.

We apply the square root equivalence scale, dividing household income by the square root of the total number of household members. This transformation accounts for economies of scale in consumption: larger households require more resources, but not proportionally more per additional member.

As a result, all individuals within a household are assigned the same equivalised income, regardless of their age or role within the household.

```{r}
data_eqv <- data_tb_coded %>%
     lissyrtools::apply_sqrt_equivalisation(
        var_name = 'dhi'
        )
```

### Poverty Line

At this point, we have done all the adjustments to original `dhi` variable, and reduced the dataset to the household-level. We are now settled to compute a poverty line that we will establish to be 50% of the weighted median of equivalised income

We can use the already familiar `run_weighted_percentiles()` that by default will compute the Minimum the Maximum and the 25th, 50th and 75th percentiles, and output themn in a named vector inside a list.

```{r,  class.output = "foldable"}
lissyrtools::run_weighted_percentiles(
        data_eqv[1], 
        var_name = "dhi", 
        wgt_name = "new_wgt"
    )
```

Therefore, we can use the `map()` function and retreive only the elemet corresponding to the Median, named "50%", and then multiply it by 0.5 to establish the poverty line. This relative definition ensures that poverty is assessed in relation to the living standards prevailing within each dataset, and thus there is no need to adjust for inflation or for diffferent currencies.

```{r}
medians <- purrr::map(
    lissyrtools::run_weighted_percentiles(
        data_eqv, 
        var_name = "dhi", 
        wgt_name = "new_wgt"
    ), "50%"
)
poverty_lines <- purrr::map(medians  , ~ .x  * 0.5) # Half of the Median

print(poverty_lines)
```

It is important to note that poverty lines are always calculated based on the full population. These same thresholds are subsequently used when evaluating poverty rates for specific subgroups, such as children or the elderly.

### Conditional Variable Construction

With the poverty line established, we create an indicator identifying whether each household falls below this threshold. For each observation, we compare equivalised income (`dhi`) to the poverty line.

Households with income below the threshold are classified as poor and assigned a value of 100, while all others are assigned 0. This binary indicator forms the basis for computing poverty rates.

```{r}
data_with_pov_line <- purrr::map2(
  data_eqv,
  poverty_lines,
  ~ .x %>%
    dplyr::mutate(
      line = .y,
      poor = dplyr::if_else(dhi < line, 100, 0)
    )
)
```

::: callout-tip
When using `purrr::map()2`, '.x' refers to the first element of the first list `data_eqv`, whilst '.y' to the first element of the second list `povery_lines`.
:::

By expressing the indicator as 0/100 rather than 0/1, the resulting weighted mean directly corresponds to a percentage.

### Weighted Aggregation

In the final step, we restrict the sample to the population of interest: children living in single-mother households. This is defined as households where the reference person is female (`sex == 2`), has no partner present (`hpartner == 0`), and where at least one child is present (`nhhmem17 > 0`).

We then compute the weighted mean of the poverty indicator using the child-specific weight (`hwgt * nhhmem17`). This produces the poverty rate among children in single-mother households.

Throughout this process, the poverty line remains fixed—having been calculated for the total population—while the weighting scheme ensures that the estimate accurately reflects the distribution of children across households.

```{r}
data_with_pov_line %>% 
  purrr::map( ~ .x %>%
         filter(hpartner==0 & sex==2 & nhhmem17>0)
  ) %>% 
  run_weighted_mean("poor", "new_chld_wgt")
```